---
title: "Getting Started with GeosptNet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with GeosptNet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

In this package we will use the South-West region of England to illustrate how the algorithm works. Several datasets are supplied with the packages:

- `BristolBathGraph` An `igraph` object representing an unidirected graph containing a set of vertexes and edges. In this graph, vertexes represent postcode sectors (e.g. *BS1 1*, *BS1 2*, etc) and edges represent transportation link between the incident vertexes. Vertexes contain an attribute called `population` which is collected from 2011 UK census data. Edges have two attributes `duration` and `distance` representing the transportation cost.
- `BristolBathPlaces` A `dataframe` object containing the locations of several towns and cities in South-West England.
- `BristolBathPolygons` A `SpatialPolygons` object of postcode sectors in the South-West region of England.

```{r setup, fig.width=5, fig.height=6, message=FALSE}
library(sp)
library(igraph)
library(GeosptNet)
BristolBathGraph
BristolBathPlaces
plot(BristolBathPolygons)
```

The `igraph` object `BristolBathGraph` contains many vertexes and edges so it is much easier to visualise a subset of it. We may use the `induced_subgraph()` function in the `igraph` pachage to select a smaller part of the graph. The graph can be easily visualised using vanilla `plot()` function or visualised using `ggplot2` and `ggnetwork` packages. The example below shows all postcode sectors within the **BA1** area.

In practice, the road distance and travel time can be retrived using commercial API (e.g. [Google Distance Matrix API](https://developers.google.com/maps/documentation/distance-matrix/overview)) or derived using [OSRM](https://github.com/riatelab/osrm).

```{r, fig.width=7, fig.height=5, message=FALSE}
library(dplyr)
library(ggplot2)
library(ggnetwork)
set.seed(1000)
BristolBathGraph %>%
  induced_subgraph(vids = which(grepl("^(BA1 )", vertex_attr(., "name")))) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(colour="grey20") +
  geom_nodes(aes(size=population), colour="grey60") +
  geom_nodetext(aes(label=name), color = "blue", size=3) +
  geom_edgetext(aes(label=sprintf("%s mins\n(%s km)", round(duration/60), round(distance/1000))),
                size = 3) +
  theme_blank()
```

Many parts of the algorithm requires using a \(N \times N\) cost matrix so this is pre-calculated and stored as an object . The example below calculates the shortest travel duration to and from any vertexes in the graph.

```{r}
my_quickest_paths <- distances(
  graph = BristolBathGraph,
  weights = edge_attr(BristolBathGraph, "duration"))

# Visualise part of the matrix
my_quickest_paths[1:10, 1:10]
```

## Detecting Communities

This package has two core functions `detect_communities()` and `merge_communities()`. Both functions accept multiple parameters with the first argument being a `data.frame` (or `tibble`) object, which means both functions are fully compatible with the pipe operator `%>%`. 

The example below shows how to split a single region (i.e. `SW England` in this case)  into multiple smaller regions. The first function call shows a simple example of using `detect_communities()` to split `l1` zones (i.e. the highest level) into multiple smaller zones at `l2`. It uses `igraph::cluster_louvain()` to determine membership of the graph communities.

The second function call show how optional parameters can be used. In particular, the `allow_exit_zone` value is set to `TRUE` which means transportation links between non adjacent vertexes may be allowed up to the length specified by the optional parameter `max_non_adjacent_path_length`. The parameter `within_zones` is set to a character vector containing names of the `l2` zones which we wish to split further. In this example, the names of the `l2` zones are dynamically calculated by the function `get_matrix_aggregate()`. This will use the matrix input `my_quickest_paths` to calculate the 95<sup>th</sup> percentile of the quickest travel duration within all `l2` zones and return only those which exceed 30 minutes threshold, only those `l2` zones satisfying the threshold will be split further down into smaller zones at `l4`.

```{r}
library(magrittr)

z <- tibble(
    name = vertex_attr(BristolBathGraph, "name"),
    l1 = "SW England") %>%
  detect_communities(
    g = BristolBathGraph,
    at_level = "l1",
    assign_level = "l2",
    edge_attribute = "duration") %>%
  detect_communities(
    g = BristolBathGraph,
    at_level = "l2",
    assign_level = "l4",
    edge_attribute = "duration",
    allow_exit_zone = TRUE,
    m = my_quickest_paths,
    max_non_adjacent_path_length = 2,
    within_zones = get_matrix_aggregate(
      g = BristolBathGraph,
      m = my_quickest_paths,
      groups = .[["l2"]],
      func = quantile,
      probs = 0.95,
      names = FALSE) %>% 
      extract(.>(60*60*0.5)) %>%
      names())

z
```

## Merging Communities

Smaller zones at low level can be merged to create larger zone at high level using the function `merge_communities()`. The algorithm will calculate the vertex aggregate value specified by parameters `vertex_attribute`, `vertex_aggregate_func`,`vertex_aggregate_lower_threshold`, `vertex_aggregate_upper_threshold` and `vertex_aggregate_args` (optional). All the zones which satisfy the conditions will be ranked and attempt to merge with one of its adjacent zone which shares the same parent zone at `parent_level`. Modularity function is used to compare which adjacent zone it should merge with. The algorithm will then calculate the cost aggregate value of the proposed zone specified by `cost_aggregate_func` and `cost_aggregate_args` (optional). The proposed zones will only be created if the cost aggregate value is within the range set by `cost_lower_threshold` and `cost_upper_threshold`.

The following code demonstrates using `merge_communities()` to merge  smaller `l4` zones into large zones at `l3`. In this example, all `l4` zones with total population between 0-75000 will attempt to merge with one of their adjacent zones. Merging will proceed if the 95<sup>th</sup> percentile travel duration of the proposed zone is between 0-45 minutes.

```{r}
z <- z %>%
  merge_communities(
    g = BristolBathGraph,
    m = my_quickest_paths,
    at_level = "l4",
    assign_level = "l3",
    vertex_attribute = "population",
    vertex_aggregate_func = sum,
    vertex_aggregate_lower_threshold = 0,
    vertex_aggregate_upper_threshold = 75000,
    edge_attribute = "duration",
    cost_aggregate_func = quantile,
    cost_lower_threshold = 0,
    cost_upper_threshold = (60*45),
    parent_level = "l2",
    cost_aggregate_args = list(probs = 0.95, 
                               names = FALSE),
    verbose = TRUE)

z
```

Let's calculate the number of distinct zones at each level:
```{r}
summarise_all(z,function(x){ length(unique(x)) })
```
## Visualisation

The `SpatialPolygons` object `BristolBathPolygons` can be merged using `maptools::unionSpatialPolygons()`. Afterwards it can be converted into `data frame` by `fortify()` and visualised using `ggplot2`.

```{r, fig.width=8, fig.height=10}
library(ggrepel)
library(maptools)

l1_polygons <- unionSpatialPolygons(
  SpP = BristolBathPolygons,
  IDs = z[match(BristolBathPolygons$Name,z$name),]$l1)
l2_polygons <- unionSpatialPolygons(
  SpP = BristolBathPolygons,
  IDs = z[match(BristolBathPolygons$Name,z$name),]$l2)
l3_polygons <- unionSpatialPolygons(
  SpP = BristolBathPolygons,
  IDs = z[match(BristolBathPolygons$Name,z$name),]$l3)
l4_polygons <- unionSpatialPolygons(
  SpP = BristolBathPolygons,
  IDs = z[match(BristolBathPolygons$Name,z$name),]$l4)

bind_rows(mutate(fortify(l1_polygons), level="l1"),
          mutate(fortify(l2_polygons), level="l2"),
          mutate(fortify(l3_polygons), level="l3"),
          mutate(fortify(l4_polygons), level="l4")) %>%
  ggplot() +
  geom_polygon(aes(x=long, y=lat, group=group, fill=id), 
               colour="black", size=0.05)+
  geom_label_repel(aes(x=long, y=lat,label=place),
                   size=3, label.size = 0.2,
                   BristolBathPlaces)+
  geom_point(aes(x=long, y=lat), BristolBathPlaces)+
  coord_map() +
  facet_wrap("level",ncol = 2) +
  scale_fill_viridis_d() +
  theme(legend.position = "none")
```

Further details of zones at a particular level can be calculated and passed onto `ggplot2` for visualisation.

```{r}
library(rgeos)

zone_centroids <- gCentroid(l3_polygons, byid = TRUE) %>%
  as.data.frame() %>%
  mutate(id=rownames(.)) %>%
  as_tibble()

zone_populations <- get_vertex_attr_aggregate(
    g = BristolBathGraph,
    attr = "population",
    groups = z[["l3"]],
    func = sum) %>% 
  tibble(id=names(.), total_pop=.)

zone_durations <- get_matrix_aggregate(
    g = BristolBathGraph,
    m = my_quickest_paths,
    groups = z[["l3"]],
    func = quantile,
    probs = 0.95,
    names = FALSE) %>% 
  tibble(id=names(.), duration_perc95=.)

zone_summary <- zone_centroids %>%
  inner_join(zone_populations, by="id") %>%
  inner_join(zone_durations, by="id") %>%
  mutate(label_text = sprintf(fmt = "%s mins\nPop: %s", round(duration_perc95/60), total_pop))

zone_summary
```

The 95<sup>th</sup> percentile travel duration and total population are visualised for all zones at `l3`.

```{r, fig.width=8, fig.height=8}
ggplot() +
  geom_polygon(aes(x=long, y=lat, group=group, fill=total_pop), 
               data = l3_polygons %>%
                        fortify() %>%
                        inner_join(zone_populations, by="id"),
               colour="black", size=0.05) +
  geom_label_repel(aes(x=x, y=y,label=label_text),
                   data = zone_summary,
                   box.padding   = 0.8, 
                   label.padding = 0.2,
                   force = 20,
                   alpha=0.8,
                   size=3, 
                   label.size = 0.2) +
  coord_map() +
  scale_fill_viridis_c() +
  labs(fill= "Zonal Population",
       x="Longitude",
       y="Latitude",
       title = "L3 Zones") +
  theme(legend.position = "right")
```
